\documentclass[11pt]{letter}



\usepackage[hmargin=2cm,vmargin=2.5cm]{geometry}



\begin{document}



\thispagestyle{empty}



Dear Professor Jermaine,



\vspace*{1cm}



Please find enclosed the manuscript ``Expressive power of linear algebra query languages", which was invited to be submitted to ACM Transaction on Database Systems as one of the best papers of PODS 2021 conference. %Please note that due to the fact that this paper is an invited PODS paper, the submission will be single blind.
We also did our best effort to make the submission double blind by using anonymous citations.


As per requirements listed on the journal website, this cover letter includes the list of authors, a conflict of interest list, and a statement on prior publication policy.


The required details are as follows:

\noindent{\bf Full list of authors:}
\begin{itemize}\itemsep=-1pt
\item Floris Geerts
\item Thomas Mu\~noz
\item Cristian Riveros
\item Domagoj Vrgo\v{c}
\end{itemize}

%\medskip
%
%\noindent{\bf Information on anonymous citation:} The only anonymous citation in the paper is listed as: {\em Anonymous. 20XX. Title removed to keep the submission double blind. In Anonymous.,} and it corresponds to the paper {\em Leonid Libkin, Juan L. Reutter, Domagoj Vrgo\v{c}. 2013. Trial for RDF: adapting graph query languages for RDF data. PODS 2013, pages 201--212}. Details of how the two papers are connected is detailed at the end of the introduction to the submitted manuscript.

\medskip

\noindent{\bf Conflicts of interest according to category:}
\begin{itemize}\itemsep=-1pt
\item {\em Known family relationship as spouse, child, sibling, or parent:} None that could review the manuscript.
\item {\em Business or professional partnership:} Wenfei Fan, Peter Buneman.
\item {\em Past or present association as thesis advisor or thesis student:} Michael Benedikt, Marcelo Arenas, Jan Van den Bussche, Leonid Libkin, Alejandro Grez.
\item {\em Collaboration within four years:} Pablo Barcel\'o, Wim Martens, Egor V. Kostylev, Renzo Angles, Aidan Hogan, Juan L. Reutter, Adri\'an Soto, Matthieu Mosser, Fernando Pieressa, Francisco Maturana, Jorge Baier, Dietrich Daroch, Pierre Bourhis, Fernando Su\'arez, Matias Junemann, Felipe Pezoa, Miguel Romero, Tony Tan, Filip Mazowiecki, Mart\'in Mu\~noz, Nathana\"el Fijalkow, James Worrell, Gabriele Puppis,  Slawek Staworko, Daniel Hern\'andez, Carlos Rojas, Enzo Zerega,  Jan Van den Bussche, Guillermo A. P\'erez, Maksimilian Ryschkov, Leopoldo E. Bertossi, Joeri Rammelaere, Giansalvatore Mecca, Paolo Papotti, Donatello Santoro, Robert Brijder, Timmy Weerwag, Maarten Van den Heuvel, Peter Ivanov, Wolfgang Gatterbauer, Martin Theobald, Bart Kuijpers, Yang Cao, Ping Lu, Fabian Pijcke, Jef Wijsen, Bart Goethals.
%Hannes Voigt, Wolfgang Lehner, Sergi Nadal, Alberto Abelló, Oscar Romero, Panos Vassiliadis, Jelle Hellings, Marc Gyssens, Yuqing Wu, Dirk Van Gucht, George H. L. Fletcher, Alexander Artikis, Alessandro Margara, Matthias Weidlich, Jonny Daenen, Frank Neven, Ronald Fagin, Benny Kimelfeld, Frederick Reiss, Erik Agterdenbos, Chee-Yong Chan, Yoav Nahshon, Liat Peterfreund, Marc Gyssens, Dirk Leinders,  Yuqing Wu, Dimitri Surinx, Andrea Cali, Davide Martinenghi, Igor Razgon, Carlos Buil Aranda, Michel Dumontier, Denis Parra, Freddie Venegas.
\item {\em Co-editing within two years:} N/A.
\item {\em Other:} Jorge Per\'ez, Claudio Gutierrez.
\end{itemize}

\medskip

\noindent{\bf Anonymous citations:} The only anonymous citation, indexed as [5] in the submission, is the original conference version of the paper published as:
\begin{itemize}
\item {\em Floris Geerts, Thomas Mu\~noz, Cristian Riveros,  Domagoj Vrgo\v{c}. 2021. 
Expressive power of linear algebra query languages. PODS 2021, pages 342--354}.
\end{itemize}

\medskip

\noindent{\bf Statement on Prior Publication Policy:} The paper {\em Floris Geerts, Thomas Mu\~noz, Cristian Riveros,  Domagoj Vrgo\v{c}. 2021. 
Expressive power of linear algebra query languages. PODS 2021, pages 342--354} falls into the category (c) of the TODS Prior Publication Policy. As discussed in the Introduction to the manuscript, the work we are submitting is a full version of the listed conference proceedings paper, and we consider it to be a major value-added extension, as required by the submission guidelines. In the Introduction to the paper we also listed the new contributions of this manuscript, and why they bring more than 30\% of new material.

\medskip


We confirm that this manuscript has not been published elsewhere and is not under consideration by another journal.

Please address all correspondence to:

Floris Geerts\\
University of Antwerp\\
Campus Middelheim\\
Middelheimlaan 1\\
2020 Antwerpen, BEL\\
floris.geerts@uantwerp.be



%Domagoj Vrgo\v{c}\\
%PUC Chile\\
%Vicuna Mackenna 4860\\
%Edificio San Agustin\\
%4to piso\\
%Macul 7820436\\
%domagojvrgoc@gmail.com\\
%dvrgoc@ing.puc.cl

\vspace*{1cm}


Best Regards,\\
The authors.
%Domagoj Vrgo\v{c}.


\end{document}



