\begin{proof}
    We show undecidability based on the following undecidable language:
    $$
    \{ \langle M\rangle\mid \text{$M$ is a deterministic TM which halts on the empty input}\},
    $$
    where $\langle M\rangle$ is some string encoding of $M$.
    Consider a TM $M$ described by $(Q,\Gamma=\{0,1\},q_0,q_m,\Delta)$
    with $Q=\{q_1,\ldots,q_m\}$ its states, $q_1$ being the initial state and $q_m$ being
    the halting state, $\Gamma$ is the tape alphabet, and $\Delta$ is a transition function
    from $Q\times \Gamma\to Q\times\Gamma\times \{\leftarrow,\sqcup,\rightarrow\}$. The simulation
    of linear space TM, as given in the proof of Proposition~\ref{prop:transducer} can be easily modified to
    any TM $M$ provided that we limit the execution of $M$ to exactly $n$ steps. Let $e_M$ denote this expression. Similarly
    as in the linear space TM simulation, we have vector variables $Q_1,\ldots,Q_m$ encoding the
    states, a single relation $T$ encoding the tape and relation $H_T$ encoding the position
    of the tape.  When an instance $\I$ assigns $n$ to $\alpha$, we have a tape of length $n$ at our disposal. This suffices if we let $M$ run for $n$ steps. We further observe that all vector variables can be assumed to be zero, initially.
    This is because we do not have input. So, let $\I_n^0$ denote the instance which assigns vector variables to the $n$-dimensional zero vector.  Furthermore, by contrast to the linear space TM simulation, we use a single vector $v$ (instead of $k$ such vectors) to simulate $n$ steps of $M$. Finally, we modify the expression given in the proof of Proposition~\ref{prop:transducer} such $\sem{e_M}{\I_n^0}$  returns $1$ if $M$
    halts in at most $n$ steps, and $0$ if $M$ did not halt yet after $n$ steps.

    As a consequence, when $M$ does not halt, $\sem{e_M}{\I_n^0}=0$ for all $n\geq 0$. When $M$ halts, there will be an $n$ such that $\sem{e_M}{\I_n^0}=1$ It now suffices to consider the \langfor expression
    $$
    d_M:=e_M\cdot e_{\mathsf{exp}}
    $$
    where $e_{\texttt{exp}} = \ffor{v}{X=\ones(X)^T\cdot\ones(X)}{X\cdot X}$ such that
    $e_{\texttt{exp}}(\I_n^0)=n^{2^n}$. Then, when $M$ does not halt we can clearly compute $d_M$ with a constant degree circuit ``0''
    for any $n$, otherwise, the circuit needed will be of exponential degree
    for at least one $n$, simply because no polynomial degree uniform  circuit family can compute $n^{2^n}$. In other words, deciding whether $d_M$ has polynomial degree coincides with deciding whether $M$ halts.
\end{proof}